---
title: "A4: Incarceration"
output: html_document
---

### Introduction

For many decades, the United States has faced increasing amounts of disparities regarding the racial demographics in jails and prisons. Stricter government policies along with increased police presence in certain regions has led to the mass incarceration of Black, Latinx, APPI, etc. people. While this affects the demographics of jails and prisons, it also affects American society. This can influence what neighborhoods are policed more often and even who is targeted by the police. To better understand this issue, I will be observing and analyzing data sets to compare the total prison populations between White and Black Americans, as well as their difference in mean prison population per 100,000 people in various counties. These factors will also be compared over time, from 1990-2010.

### Summary Information

When investigating these data sets, the first value I wanted to find was the difference between the total White prison population and total Black prison population in 2010. This would give us the broadest overview of any racial disparities in the total prison population. The prison population of White Americans was **448,158** and for Black Americans it was **488,969**. This tells us that there is are **40,811** more Black Americans in prison than White Americans. While this gives us an idea of some sort of difference in the total prison population, the prison population *rate* is more telling of racial disparities due to differences in racial populations in the United States. I chose to find values that would compare the mean prison rate per 100,000 people across all counties between White and Black Americans, and then I chose to compare those numbers from 1990 to 2010. In 1990 White Americans had an average of **249.42** people in prison per 100,000 people, while Black Americans had an average of **2028.65** people in prison per 100,000 people. Immediately you can notice there is a large gap between the two groups. Jumping forward, I also calculated those statistics 20 years later in 2010. White Americans had a rate of **528.76**, while Black Americans had a rate of **3237.458**. This shows that even 2 decades later, the disparity still remains. 

### The Dataset

The data used in this report was collected by the **Vera Institute**. According to their website, they do research and collect this data by working alongside community members and local organizations and government. They say that their mission is to is to end the overcriminalization and mass incarceration of people of color, immigrants, and people experiencing poverty. They do this work in order to develop an idea for how to disrupt the incarceration system and its devastating impacts. In the prison population and rate datasets there are **153,811 observations**. The total prison population dataset has **36 features** and the prison population rate dataset has **23 features**. Because this dataset deals with large numbers and widespread information regarding a historically systematic topic like race, there are some questions of power that need to be considered. For example, it is important to consider the identities of the individuals who collected this data as their own personal perspectives could affect what they did/didn't include. Additionally, one may want to consider who has control over this data and what may have been withheld or not reported as this can influence the accuracy of the reflections of the populations we are observing. This connects to limitations the data may have, including the many cells of unreported or "NA" data. This can cause issues with data accuracy and drawing valid conclusions from the data.

### Trends Over Time Chart

This chart was designed to display a visual representation of the growth of the prison population since 1990. The color coded lines also indicate the Black and White proportional makeup of the prison population in that year. One prominent pattern that emerged was the significant growth of the overall prison population after 1991. Since then it has steadily risen and fallen. Another pattern that you can see regards the racial demographics of the population. Although both populations have grown over time, the consistent majority dominating the prison populations is Black Americans. This displays disparities in incarceration when it comes to race.

```{r, echo = FALSE, code = readLines("time_trend.R")}
source("time_trend.R")
```

### Variable Comparison Chart

I included this chart to highlight disparities in Black and White Prison Populations in 2010 and visually display that data. Each dot is a location from the data set and it is placed on the graph wherever its corresponding Black and White prison population is. In a perfect world, the dots on the plot would be in a straight proportionate line from the bottom left to the top right. However, you can see on this plot when you study the scale and positioning of each dot that a majority of the points indicate a higher x-value (Black prison population) than y-value (White prison population). This shallowness of this plots' trend shows a disproportionate number of incarcerated Black Americans in an overwhelming majority of cities.

```{r, echo = FALSE, code = readLines("variable_comparison.R")}
source("variable_comparison.R")
```

### Map

I included this map to create a easy-to-read visual that helps geographically gauge various concentrations of Black prison population rates in the United States in 2010.

```{r, echo = FALSE, code = readLines("map.R")}
source("map.R")
```